---
title: "Mapping Occupation and Race in Boston During COVID-19"
author: "Lauren Chambers"
date: "5/1/2020"
output:
  html_document:
    df_print: paged
---

# Getting Started
```{r, results="hide", message=FALSE, warning=FALSE}
# Import needed libraries
library(tidyverse)
library(leaflet)
library(tidycensus)
library(rgdal)
library(mapview)
library(readxl)

# Set ggplot theme
theme_set(theme_minimal())
```


## Examine ACS tables
Explore the variables available in the census data
```{r}
# Get all ACS 2018 subject tables & columns names
all_acs18_subj_tabs <- tidycensus::load_variables(2018, "acs5/subject", cache = TRUE)
View(all_acs18_subj_tabs)

# Get all ACS 2018 tables & columns names
all_acs18_tabs <- tidycensus::load_variables(2018, "acs5", cache = TRUE)
View(all_acs18_tabs)
```

## Define helper functions for plotting
Do math to figure out percentage of estimate variables
```{r}
get_percents <- function(df, save=F) {
  
  # Get the input data frame's name as a string
  original_df_name <- deparse(substitute(df))
  
  # Sum together the various "estimate" values and calculate percent of the whole
  df <- df %>%
    group_by(GEOID) %>%
    mutate(sum_value = sum(estimate),
           sum_moe = sum(moe),
           percent = sum_value / summary_est * 100) %>%
    select(-estimate, -moe, -variable) %>%
    unique() %>%
    # Exclude census tracts with no land
    filter(ALAND > 0)
  
  # If desired, save resulting dataframe & spatial data to file
  if (save) {
    writeOGR(obj=df, dsn=paste0("app/data/", original_df_name), 
             layer=original_df_name, driver="ESRI Shapefile")
  }
  
  return(df)
}
```

Replace legend labels with custom format
```{r}
myLabelFormat = function(..., custom_percent=F, income=F){ 
  # Add "≥" in front of last legend percentage to denote lower limit
  if(custom_percent){ 
    function(type = "numeric", cuts){ 
      cuts <- paste0(cuts, "%")
      cuts[[length(cuts)]] <- paste0("≥", cuts[[length(cuts)]])
      cuts
    } 
    
  # Add "≥" in front of last legend value to denote lower limit & add 
  # commas for legibility in big numbers
  } else if(income){ 
    function(type = "numeric", cuts){ 
      cuts <- paste0("$",format(cuts ,big.mark=",",scientific=FALSE))
      cuts[[length(cuts)]] <- paste0("≥", cuts[[length(cuts)]])
      cuts
    } 
  
  # Use default label formatter  
  } else{
    labelFormat(...)
  }
}
```

# Get internet/computer access by census tract
## Download data
```{r}
# Download ACS census data on internet/computer access:
# - B28008_001: Total population in tract 
# - B28008_009: Population in tract with computer but no internet
# - B28008_010: Population in tract with no computer
ma_intcomp_tract <- get_acs(geography = "tract",
                            state="MA",
                            variables = c("B28008_009", "B28008_010"),
                            summary_var = "B28008_001", 
                            keep_geo_vars = T,
                            geometry=T)
```

## Define mapping data & parameters
```{r}
# Calculate percent of no computer + computer w/o internet
to_plot <- get_percents(ma_intcomp_tract)

# Save to shapefile
writeOGR(obj=to_plot, dsn=paste0("app/data/", "intcomp_by_tract"), 
             layer="intcomp_by_tract", driver="ESRI Shapefile")

# Define percent values to plot
domain_to_plot <- 0:30

# Define color palette for partial range of percentages
pal_intcomp <- colorNumeric(
  palette = "viridis",
  domain = domain_to_plot,
  na.color = "#FDEF23" # yellow at end of viridis
)
```

## Create map
```{r}
# Create interactive Leaflet map
m <- leaflet() %>%
  # Use gray background map
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  # Add census tracts colored by data
  addPolygons(data = to_plot, 
              fillColor = ~pal_intcomp(percent), # Fill color base on percentage
              color = "none",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.7,
              group="poly") %>%
  # Add legend
  addLegend(pal = pal_intcomp, 
            values = domain_to_plot,
            position = "topright", 
            title = "<a style='font-family:GT America; color: dimgrey'>
                        Percent of <br>residents without<br>computer or internet<br>by census tract
                    </a>",
            labFormat = myLabelFormat(custom_percent=T)
            )  %>%
  # Add button to reset zoom
  addEasyButton(easyButton(
    icon="fa-home", title="Reset",
    onClick=JS("function(btn, map){ 
                   var groupLayer = map.layerManager.getLayerGroup('poly');
                   map.fitBounds(groupLayer.getBounds());
               }")))

# Show map
m
```

## Save map as PNG
```{r}
paste0(as.character(getwd()), "/figures/",
       strftime(Sys.time(), "%Y%m%d_%H%M%S"),
       "_tracts_30pct.png") %>%
  mapshot(m, file = .,
          vwidth=700, vheight=500, expand=2,
          remove_controls = c("zoomControl", "homeButton"))
```

# Get broadband internet access by census tract
## Download data
```{r}
# Download ACS census data on broadband access:
# - B28008_001: Total population in tract 
# - B28008_005: Population in tract with fixed broadband internet
ma_broadband_tract <- get_acs(geography = "tract",
                              state="MA",
                              variables = "B28008_005",
                              summary_var = "B28008_001", 
                              keep_geo_vars = T,
                              geometry=T)
```

## Define mapping data & parameters
```{r}
# Calculate percent of non-broadband households
to_plot <- ma_broadband_tract %>%
  select(-moe, -summary_moe) %>%
  rename(total = summary_est,
         broadband = estimate) %>%
  # Calculate non-broadband
  mutate(non_broadband = total - broadband) %>%
  # Calculate percentage
  mutate(percent = 100 * (non_broadband / total)) %>%
  filter(!is.na(percent)) 

# Save to shapefile
writeOGR(obj=to_plot, dsn=paste0("app/data/", "broadband_by_tract"), 
             layer="broadband_by_tract", driver="ESRI Shapefile")

# Define percent values to plot
domain_to_plot <- 0:30

# Define color palette for partial range of percentages
pal_broadband <- colorNumeric(
  palette = "viridis",
  domain = domain_to_plot,
  na.color = "#FDEF23" # yellow at end of viridis
)
```

## Create map
```{r}
# Create interactive Leaflet map
m <- leaflet() %>%
  # Use gray background map
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  # Add census tracts colored by data
  addPolygons(data = to_plot, 
              fillColor = ~pal_broadband(percent), # Fill color base on percentage
              color = "none",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.7,
              group="poly") %>%
  # Add legend
  addLegend(pal = pal_broadband, 
            values = domain_to_plot,
            position = "topright", 
            title = "<a style='font-family:GT America; color: dimgrey'>
                        Percent of residents<br>without broadband<br>by census tract
                    </a>",
            labFormat = myLabelFormat(custom_percent=T)
            )  %>%
  # Add button to reset zoom
  addEasyButton(easyButton(
    icon="fa-home", title="Reset",
    onClick=JS("function(btn, map){ 
                   var groupLayer = map.layerManager.getLayerGroup('poly');
                   map.fitBounds(groupLayer.getBounds());
               }")))

# Show map
m
```

## Save map as PNG
```{r}
paste0(as.character(getwd()), "/figures/",
       strftime(Sys.time(), "%Y%m%d_%H%M%S"),
       "_broadband_tracts_30pct.png") %>%
  mapshot(m, file = .,
          vwidth=700, vheight=500, expand=2,
          remove_controls = c("zoomControl", "homeButton"))
```

# Get income by census tract
## Download data
```{r}
# Download ACS census data on income:
# - S1902_C03_001: Mean household income
ma_income_tract <- get_acs(geography = "tract",
                            state="MA",
                            variables = "S1902_C03_001",
                            keep_geo_vars = T,
                            geometry=T)
```

## Define mapping data & parameters
```{r}
# Prep data for mapping
to_plot <- ma_income_tract %>%
  rename(income = estimate) %>% 
  filter(!is.na(income))

# Save to shapefile
writeOGR(obj=to_plot, dsn=paste0("app/data/", "income_by_tract"), 
             layer="income_by_tract", driver="ESRI Shapefile")

# Define income values to plot
domain_to_plot <- 20000:140000

# Define color palette for partial range of incomes
pal_income <- colorNumeric(
  palette = "viridis",
  na.color = "#3D0C51", # Start of viridis scale
  reverse=T, # Low income colors should match low internet colors
  domain = domain_to_plot
)
```

## Create map
```{r}
# Create interactive Leaflet map
m <- leaflet() %>%
  # Use gray background map
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  # Add census tracts colored by data
  addPolygons(data = to_plot, 
              fillColor = ~pal_income(income), # Fill color base on income
              color = "none",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.7,
              group="poly") %>%
  # Add legend
  addLegend(pal = pal_income, 
            values = domain_to_plot,
            position = "topright", 
            title = "<a style='font-family:GT America; color: dimgrey'>
                        Mean household income<br>by census tract
                    </a>",
            labFormat = myLabelFormat(income=T)
            )  %>%
  # Add button to reset zoom
  addEasyButton(easyButton(
    icon="fa-home", title="Reset",
    onClick=JS("function(btn, map){ 
                   var groupLayer = map.layerManager.getLayerGroup('poly');
                   map.fitBounds(groupLayer.getBounds());
               }")))

# Show map
m
```

## Save map as PNG
```{r}
paste0(as.character(getwd()), "/figures/",
       strftime(Sys.time(), "%Y%m%d_%H%M%S"),
       "_income_tracts_150k.png") %>%
  mapshot(m, file = .,
          vwidth=700, vheight=500, expand=2,
          remove_controls = c("zoomControl", "homeButton"))
```

# Get home language by tract
## Download data
```{r}
# Download ACS census data on language:
# - S1601_C02_003: Percent of tract speaking language other than English at home
ma_lang_tract <- get_acs(geography = "tract",
                            state="MA",
                            variables = "S1601_C02_003",
                            keep_geo_vars = T,
                            geometry=T)
```

## Define mapping data & parameters
```{r}
# Prep data for mapping
to_plot <- ma_lang_tract %>%
  rename(percent = estimate) %>%
  filter(!is.na(percent))

# Save to shapefile
writeOGR(obj=to_plot, dsn=paste0("app/data/", "lang_by_tract"), 
             layer="lang_by_tract", driver="ESRI Shapefile")

# Define percent values to plot
domain_to_plot <- 0:70

# Define color palette for partial range of percentages
pal_lang <- colorNumeric(
  palette = "viridis",
  domain = domain_to_plot
)
```

## Create map
```{r}
# Create interactive Leaflet map
m <- leaflet() %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addPolygons(data = to_plot, 
              fillColor = ~pal_lang(percent), # Fill color base on percentage
              color = "none", 
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.7, 
              group="poly") %>%
  # Add legend
  addLegend(pal = pal_lang, 
            values = domain_to_plot,
            position = "topright",
            title = "<a style='font-family:GT America; color: dimgrey'>
                        Percent of residents<br>speaking language other<br>than English at home,<br>by census tract
                    </a>",
            labFormat = myLabelFormat(custom_percent=T)
            )  %>%
  # Add button to reset zoom
  addEasyButton(easyButton(
    icon="fa-home", title="Reset",
    onClick=JS("function(btn, map){ 
                   var groupLayer = map.layerManager.getLayerGroup('poly');
                   map.fitBounds(groupLayer.getBounds());
               }")))

# Show map
m
```

## Save map as PNG
```{r}
paste0(as.character(getwd()), "/figures/",
       strftime(Sys.time(), "%Y%m%d_%H%M%S"),
       "_language_70pct_tracts.png") %>%
  mapshot(m, file = .,
          vwidth=700, vheight=500, expand=2,
          remove_controls = c("zoomControl", "homeButton"))
```

